The Unicode Cookbook \linebreak for Linguists

### 2017-01-20 book/chapters/implementation.tex
cleanup

# make softlinks to the R executables in /usr/local/bin
ln -is $pathT $pathW /usr/local/bin
@

@@

<<>>=
test <- c('AABB', '\u0041\u0410\u0042\u0412')
write.profile(test)
@

@@

Further, the current R-implementation has a limitation when regular expressions
are used. The problem is that overlapping matches are not captured when using
regular expressions.\footnote{This restriction is an effect of the underlyingly
used ICU implementation of the Unicode standard as implemented in R through the
package \texttt{stringi}.} Everything works as expected without regular
expressions, but there might be warnings/errors in case of \texttt{regex =
TRUE}. However, just as in the previous issue, this problem should only very
rarely (when at all) happen in natural language data.

The problem can be exemplified by a sequence <bbbb> in which a 
grapheme <bb> should be matched. With the default \texttt{regex = FALSE} 
there are three possible matches, but with \texttt{regex = TRUE} only the first 
two or the last two are matched. The middle two are not matched because they 
overlap with the other matches. In the example below this leads to an error, 
because the second <bb> is not matched. We have not been able to produce a real 
example in any natural language in which this limitation might lead to an error.

>>>>

# make softlinks to the R executables in /usr/local/bin
# you will have to enter your user's password!
sudo ln -is $pathT $pathW /usr/local/bin
@

@@

<<>>=
( test <- c('AABB', '\u0041\u0410\u0042\u0412') )
write.profile(test)
@

@@

Further, the current R-implementation has a limitation when regular expressions
are used. The problem is that overlapping matches are not captured when using
regular expressions.\footnote{This restriction is an effect of the underlyingly
used ICU implementation of the Unicode standard as implemented in R through the
package \textttf{stringi}.} Everything works as expected without regular
expressions, but there might be warnings/errors in case of \texttt{regex =
TRUE}. However, just as in the previous issue, this problem should only very
rarely (when at all) happen in natural language data.

The problem can be exemplified by a sequence <bbbb> in which a grapheme <bb>
should be matched. With the default \texttt{regex = FALSE} there are three
possible matches, but with \texttt{regex = TRUE} only the first two <b>s or the
last two <b>s are matched. The middle two <b>s are not matched because they
overlap with the other matches. In the example below this leads to an error,
because the second <bb> is not matched. However, we have not been able to
produce a real example in any natural language in which this limitation might
lead to an error.

### 2017-01-20 book/chapters/introduction.tex
cleanup

The necessity to computationally use and compare different orthographies most
forcefully arises in the context of language comparison. Concretely, the proper
processing of orthographies and transcriptions becomes crucial for the
development of quantitative methods for language comparison and historical
reconstruction in order to investigate worldwide linguistic variation and to
model the historical and areal processes that underlie linguistic
diversity.\footnote{
cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}}
In this work, it is crucial to be able to flexibly process across numerous
resources with different orthographies. In many cases even different resources
on the same language use different orthographic conventions. Another
orthographic challenge that we encounter regularly in our linguistic practice is
electronic resources on a particular language that claim to follow a specific
orthographic convention (often a resource-specific convention), but on closer
inspection such resources are almost always not consistently encoded. Thus a
second goal of our orthography profiles is to allow for an easy specification of
orthographic conventions, and use such profiles to check consistency and to
report errors to be corrected.

@@

Before we deal with these proposals we will extensively discuss the theoretical
background on text encoding, on the Unicode standard, and on the International
Phonetic Alphabet. In the remainder of this chapter, we give an extended
introduction to the notion of encoding (Section~\ref{encoding}) and writing
systems, both from a linguistic perspective
(Section~\ref{linguistic-terminology}) and from the perspective of the Unicode
Consortium (Section~\ref{the-unicode-approach}). We consider the Unicode
Standard to be a breakthrough (and ongoing) development that fundamentally
changed the way we look at writing systems, and we aim to provide here a
slightly more in-depth survey of the many techniques that are available in the
standard. A good appreciation for the solutions that the Unicode Standard also
allows for a thorough understanding of possible pitfalls that one might
encounter when using it (Chapter~\ref{unicode-pitfalls}). As an example of the
current state-of-the-art, we discuss the rather problematic marriage of the
International Phonetic Alphabet (IPA) with the Unicode Standard (Chapter
\ref{ipa-meets-unicode}).

The second part of the book (Sections~\ref{orthography-profiles} and
\ref{use-cases}) describes our proposals for how to deal with the Unicode
Standard in the daily practice of (comparative) linguistics. First, we discuss the
challenges of characterizing a writing system. To solve these problems, we
propose the notions of orthography profiles, closely related to Unicode locale
descriptions (Chapter~\ref{orthography-profiles}). Finally, we discuss practical
issues with actual examples (Chapter~\ref{use-cases}). We provide reference
implementation of our proposals in R and in Python, available as open-source
libraries.

@@

There are many in-depth histories of the origin and development of writing
systems \citep[e.g.~][]{Robinson1995,Powell2012}, a story that we therefore will
not repeat here. However, the history of turning writing into computer readable
code is not so often told, so we decided to offer a short survey of the major
developments of such encoding here.\footnote{Because of the recent
history as summarized in this section, we have used mostly rather ephemeral
internet sources. When not references by traditional literature in the
bibliography, we have used \url{http://www.unicode.org/history/} and various
Wikipedia pages for the information presented here. A useful survey of the
historical development of the physical hardware of telegraphy and
telecommunication is \citet{Huurdeman2003}. Most books that discuss the
development of encoding of telegraphic communication focus of cryptography,
e.g.~\citet{Singh1999}, and forego the rather interesting story of ``open'',
i.e.~non-cryptographic, encoding that is related here.} This history turns out
to be intimately related to the history of telegraphic communication.

>>>>

The necessity to computationally use and compare different orthographies most
forcefully arises in the context of language comparison. Concretely, the proper
processing of orthographies and transcriptions becomes crucial for the
development of quantitative methods for language comparison and historical
reconstruction in order to investigate worldwide linguistic variation and to
model the historical and areal processes that underlie linguistic
diversity. In this work, it is crucial to be able to flexibly process across numerous
resources with different orthographies. In many cases even different resources
on the same language use different orthographic conventions. Another
orthographic challenge that we encounter regularly in our linguistic practice is
electronic resources on a particular language that claim to follow a specific
orthographic convention (often a resource-specific convention), but on closer
inspection such resources are almost always not consistently encoded. Thus a
second goal of our orthography profiles is to allow for an easy specification of
orthographic conventions, and use such profiles to check consistency and to
report errors to be corrected.

% \footnote{cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}}

@@

Before we deal with these proposals we will first discuss the theoretical
background on text encoding, on the Unicode standard, and on the International
Phonetic Alphabet. In the remainder of this chapter, we give an extended
introduction to the notion of encoding (Section~\ref{encoding}) and the
principles of writing systems, both from a linguistic perspective
(Section~\ref{linguistic-terminology}) and from the perspective of the Unicode
Consortium (Section~\ref{the-unicode-approach}). We consider the Unicode
Standard to be a breakthrough (and ongoing) development that fundamentally
changed the way we look at writing systems, and we aim to provide here a
slightly more in-depth survey of the many techniques that are available in the
standard. A good appreciation for the solutions that the Unicode Standard
provides, also allows for a thorough understanding of possible pitfalls that one
might encounter when using it (Chapter~\ref{unicode-pitfalls}). As an example of
the current state-of-the-art, we discuss the International Phonetic Alphabet
(IPA, Chapter~\ref{the-international-phonetic-alphabet}) and its rather
problematic marriage with the Unicode Standard (Chapter
\ref{ipa-meets-unicode}).

The second part of the book describes our proposals for how to deal with the
Unicode Standard in the daily practice of linguistics. First, we discuss the
challenges of characterizing a writing system. To solve these problems, we
propose the notions of orthography profiles, closely related to Unicode locale
descriptions (Chapter~\ref{orthography-profiles}). We provide reference
implementation of our proposals in R and in Python, available as open-source
libraries. An introduction to these implementations of our proposals is
described in Chapter~\ref{implementation}. Finally, we discuss practical issues
with actual examples in Chapter~\ref{use-cases}.

@@

There are many in-depth histories of the origin and development of writing
systems \citep[e.g.~][]{Robinson1995,Powell2012}, a story that we therefore will
not repeat here. However, the history of turning writing into computer readable
code is not so often told, so we decided to offer a short survey of the major
developments of such encoding here.\footnote{Because of the recent
history as summarized in this section, we have used mostly rather ephemeral
internet sources. When not referenced by traditional literature in the
bibliography, we have used \url{http://www.unicode.org/history/} and various
Wikipedia pages for the information presented here. A useful survey of the
historical development of the physical hardware of telegraphy and
telecommunication is \citet{Huurdeman2003}. Most books that discuss the
development of encoding of telegraphic communication focus of cryptography,
e.g.~\citet{Singh1999}, and forego the rather interesting story of ``open'',
i.e.~non-cryptographic, encoding that is related here.} This history turns out
to be intimately related to the history of telegraphic communication.

### 2017-01-20 book/chapters/ipa_background.tex
cleanup

Several styles of transcription with the IPA are possible, as illustrated in the
\textit{Handbook}, and they are all valid.\footnote{For illustration of
different possibilities, see the 29 languages and their transcriptions in the
original and initial \textit{Illustrations of the IPA}
\citep[41--154]{IPA2007}} Therefore, there are different but equivalent
transcriptions, or as \citet[64]{Ladefoged1990a} captures it, ``perhaps now that
the Association has been explicit in its eclectic approach, outsiders to the
Association will no longer speak of \textit{the} IPA transcription of a given
phenomenon, as if there were only one approved style.'' Clearly not all
phoneticians agree (or will likely ever agree of course) on all aspects of the
IPA or on transcription practices. As noted above, there have been several
revisions in the IPA's long history, but the current version (2005) is
strikingly similar to the 1926 version, showing the viability of the IPA as a
common standard for linguistic transcription.

@@

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single ``correct'' transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood. Basically, the IPA allows for both a `narrow' phonetic transcription and a
`broad' phonological transcription. A `narrow' phonetic transcription may freely
use the all symbols in the IPA chart with direct reference to their phonetic
value, i.e. the transcriber can indicate with the symbols <ŋ͡m> that the
phonetic value of the attested sound is a simultaneous labial and velar closure
which is voiced and contains nasal airflow, independently of the phonemic status
of this sound in the language in question.

@@

In contrast, the basic goal of a phonemic ``broad'' transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar area. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεauɔ].
Similarly, the Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

@@

\citep[164]{IPA1999}.}

@@

\begin(comment)

>>>>

Several styles of transcription with the IPA are possible, as illustrated in the
\textit{Handbook}, and they are all valid.\footnote{For illustration of
different possibilities, see the 29 languages and their transcriptions in the
original and initial \textit{Illustrations of the IPA}
\citep[41--154]{IPA2007}} Therefore, there are different but equivalent
transcriptions, or as \citet[64]{Ladefoged1990a} captures it, ``perhaps now that
the Association has been explicit in its eclectic approach, outsiders to the
Association will no longer speak of \textit{the} IPA transcription of a given
phenomenon, as if there were only one approved style.'' Clearly not all
phoneticians agree (or will likely ever agree) on all aspects of the
IPA or on transcription practices. As noted above, there have been several
revisions in the IPA's long history, but the current version (2005) is
strikingly similar to the 1926 version, showing the viability of the IPA as a
common standard for linguistic transcription.

@@

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single ``correct'' transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood. Specifically, the IPA allows for both
a `narrow' phonetic transcription and a `broad' phonological transcription. A
`narrow' phonetic transcription may freely use the all symbols in the IPA chart
with direct reference to their phonetic value, i.e. the transcriber can indicate
with the symbols <ŋ͡m> that the phonetic value of the attested sound is a
simultaneous labial and velar closure which is voiced and contains nasal
airflow, independently of the phonemic status of this sound in the language in
question.

@@

In contrast, the basic goal of a phonemic ``broad'' transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar area. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
Similarly, the Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

@@

\citep[164]{IPA1999}.

@@

\begin{comment}

### 2017-01-20 book/chapters/preface.tex
adds notes
cleanup

% Early phoneticians: Richard Lepsius, Isaac Pitman, Alexander J. Ellis, and Alexander Melville Bell
% Later ones: Henry Sweet, L'abbé Jean-Pierre Rousselot, Eduard Sievers, Paul Passy, and Jens Otto Harry Jesperson

@@

\begin{quote}
``But as they observed articulation more and more closely, and as they took more and more languages into consideration, the number of sounds they could discriminate and the number of symbols needed to denote them both increased seemingly without limit. In fact, as early as 1848, Ellis asserted that in principle the number of different sounds is endless. With the development of instrumental phonetics, largely by L'abbe Rousselt, it soon turned out that, given sufficiently accurate measuring devices, that is in fact the case.''
\end{quote}

>>>>

% Very early attmept at notion for linguistic prosody: Joshua Steele (Prosodia Rationalis, 1779)

% Early phoneticians: Richard Lepsius (standard alphabet 1854, 1855), Isaac Pitman (Stenographic Soundhand 1837, Phonotypy 1844, English Phonotypic Alphabet 1845 with Ellis), Alexander J. Ellis (English Phonotypic Alphabet 1845), and Alexander Melville Bell (visibile speech 1864, 1867)

% Later phoneticians: Henry Sweet (Romic alphabet, aka Romic reform, in A Handbook of Phonetics 1877 - later becomes basis for IPA), L'abbé Jean-Pierre Rousselot ("founder of experimental phonetics"), Eduard Sievers, Paul Passy (founded the Phonetic Teachers' Association in 1886, later became IPAssociation), and Jens Otto Harry Jesperson

% Instrumental measurements: Ludimar Hermann (used the phonograph to test theories of vowel production; coined formant)

@@

\begin{quote}

But as they observed articulation more and more closely, and as
they took more and more languages into consideration, the number of sounds they
could discriminate and the number of symbols needed to denote them both
increased seemingly without limit. In fact, as early as 1848, Ellis asserted
that in principle the number of different sounds is endless. With the
development of instrumental phonetics, largely by L'Abbé Rousselot, it soon
turned out that, given sufficiently accurate measuring devices, that is in fact
the case.

\end{quote}
